\chapter{Extra credit problems}

The book of L\'aszl\'o Lov\'asz \cite{lovasz} is a good source of advanced exercises in graph theory.

\begin{enumerate}
\item Assume that the sequence $d_1,\dots, d_p$ is graphic,
$d_i\ge 1$ for each $i$ and 
\[d_1+\dots+d_p\ge 2\cdot(p-1).\]
Show that there is a connected graph $G$ with the degree sequence $d_1,\dots, d_p$.
\item Assume $d_1,\dots, d_p$ is a sequence of integers in a nonincreasing order.
Show that it is multigraphic if and only if $d_p\ge0$, the sum $d_1+\dots+ d_p$ is even and 
\[d_1\le d_2+\dots+d_p.\]
(A sequence of integers $d_1,\dots, d_p$ is called \textit{multigraphic} if it appears as a sequence of degrees of a multigraph.)
\item Show that in any connected graph $G$ there is a vertex $v$ such that $G-v$ is connected.
\item Let $G$ be a critical graph of chromatic number $k+1$.
Show after removing any $k-1$ edges from $G$  
the obtained graph remains connected. 
\item Assume both sequences $d_1,\dots, d_p$
 and $d_1-1,\dots, d_p-1$ are graphic.
Show that there is a graph with a 1-factor and with the degree sequence $d_1,\dots, d_p$.
\item Let $M$ be a maximal planar map.
Assume the vertcies of $M$ are colored in 3 colors.
Show that the number or regions which get all three colors is even.
\item Show that any $4$-regular graph has a $2$-factor.
\item Show that any edge of cubic graph lies in an even number of Hamiltonian cycles.
\item Let $G$ be a  connected graph.
Show that any two paths of maximum length in $G$ have a common vertex.
\item Assume that the edges of a complete graph are colored in two colors.
Show that there is a Hamiltonian cycle which either monochromatic or consists of two monochromatic paths.
\item Assume two trees $R$ and $S$ 
have the vertexes $r_1,\dots,r_n$ and $s_1,\z\dots,s_n$ correspondingly.
Assume that $R-r_i$ is isomorphic to $S-s_i$ for each $i$.
Show that $R$ is isomorphic to $S$.
\item Assume that $G$ is $3$-regular connected graph such that 
for any two vertexes $v,w$ of $G$ there is an isomorphism
$G\to G$ which sends $v$ to $w$.
Prove that $G$ remains connected after removing any $2$ edges.
\item  Let $G$ be a connected graph.
Given any two vertexes $v,w$ in $G$, denote by $d(v,w)$ the length of a shortest path containing $v$ to~$w$. 
Assume that $G$ has no triangles and
\[d(x,y)+d(v,w)=\max\{\,d(x,v)+d(y,w),d(x,w)+d(y,v)\,\}\]
for any 4 vertexes $x,y,v,w$ in $G$.
Show that $G$ is a tree.
\item Suppose $p=r(m,n)$ is the Ramsey number for $m$ and $n$%;
%that is, $p$ is the minimal integer such that any edge-coloring of the
%complete graph with $K_p$ using red and blue must contain a red $K_m$ or a blue $K_n$
.

Assume that $G$ results from $K_p$ by deleting a single edge.
Show that $G$ has a red/blue edge coloring with no red
$K_m$ and no blue $K_n$.
\item Let $T$ be a spanning tree in a graph $G$ with weighted edges.
Show that $T$ has minimal total weight if and only if the weight of any edge $(a,b)$ in $G$ is at least as large as the weight of any edge on the path from $a$ to $b$ in $T$.
\item In a group of people, for some fixed $s$ and any $k$,
any $k$ girls like at least $k-s$ boys in total.
Show that then all but $s$ girls may get married on the boys they like.
\item Let $M$ be a connected planar map and $G$ be its dual.
Show that $M$ and $G$ have the same number of spanning trees.
\item Show that the number of subgraphs of $K_{n+1}$ with all even degrees equals to the number of all subgraphs of $K_n$.

\item Find the number of spanning trees of $K_n$ which include the given edge.

\end{enumerate}

\chapter{Videos}

\begin{itemize}
 \item Huffman Encoding on LeiosOS\\ \texttt{https://www.youtube.com/watch?v=wHyUxTc2Ohk}
\end{itemize}
